### 2014-08-05 doc/thesis/typedlua.tex
Formatting the subtyping and consistent-subtyping rules

\section{Statistics about the usage of Lua}

@@

\section{Typed Lua}

\subsection{Type Language}

@@

Figure \ref{fig:typelang} presents the abstract syntax of the
Typed Lua types.
We split Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
We use first-level types to represent Lua values, and
second-level types to type expression lists, multiple assignments,
and function applications.
We will all Typed Lua types along this section.

@@

\section{Related Lua projects}

>>>>

\section{Statistics about the usage of Lua}
\label{sec:statistics}

@@

\section{Type Language}
\label{sec:types}

@@

Figure \ref{fig:typelang} presents the abstract syntax of the
Typed Lua types.
We split Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
We use first-level types to represent Lua values and
second-level types to type expression lists, multiple assignments,
and function applications.
We will explain all Typed Lua types in more detail along this section.

\section{Typed Lua through examples}
\label{sec:examples}

@@

\section{The design of Typed Lua}
\label{sec:design}

\begin{figure}[!ht]
\textbf{Subtyping}\\
\dstart
\begin{footnotesize}
$$
\begin{array}{c}
\begin{array}{c}
\mylabel{S-LITERAL}\\
\senv \vdash L \subtype L
\end{array}
\;
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-NUMBER}\\
\senv \vdash {<}{\it number}{>} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {<}{\it string}{>} \subtype \String
\end{array}
\;
\begin{array}{c}
\mylabel{S-BASE}\\
\senv \vdash B \subtype B
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-NIL}\\
\senv \vdash \Nil \subtype \Nil
\end{array}
\;
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash T \subtype \Value
\end{array}
\;
\begin{array}{c}
\mylabel{S-ANY}\\
\senv \vdash \Any \subtype \Any
\end{array}
\;
\begin{array}{c}
\mylabel{S-SELF}\\
\senv \vdash \Self \subtype \Self
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
       \senv \vdash T_{2} \subtype T}
      {\senv \vdash T_{1} \cup T_{2} \subtype T}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash T \subtype T_{1}}
      {\senv \vdash T \subtype T_{1} \cup T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash T \subtype T_{2}}
      {\senv \vdash T \subtype T_{1} \cup T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FUNCTION}\\
\dfrac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
       \senv \vdash R_{1} \subtype R_{2}}
      {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash F_{j} \subtype F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
       \senv \vdash T_{1} \subtype T_{2} \;\;\;
       \senv \vdash T_{2} \subtype T_{1}}
      {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
       \senv \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
       \senv \vdash T_{1} \subtype T_{2}}
      {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARIABLE}\\
\dfrac{X_{1} \subtype X_{2} \in \senv}
      {\senv \vdash X_{1} \subtype X_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-RECURSIVE1}\\
\dfrac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-RECURSIVE2}\\
\dfrac{X \subtype X \not\in \senv \;\;\;
      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-RECURSIVE3}\\
\dfrac{X \subtype X \in \senv \;\;\;
       \senv \vdash X \subtype T_{2}}
      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-RECURSIVE4}\\
\dfrac{X \subtype X \not\in \senv \;\;\;
       \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-RECURSIVE5}\\
\dfrac{X \subtype X \in \senv \;\;\;
       \senv \vdash T_{1} \subtype X}
      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VOID}\\
\senv \vdash \Void \subtype \Void
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
      {\senv \vdash T_{1}* \subtype T_{2}*}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE1}\\
\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
       \senv \vdash S_{1} \subtype S_{2}}
      {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TUPLE2}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
       \senv \vdash T_{1}* \subtype S_{2}}
      {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE3}\\
\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
       \senv \vdash S_{1} \subtype T_{2}*}
      {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
\end{array}
\end{array}
$$
\end{footnotesize}
\dend
\caption{The subtyping relation}
\label{fig:subtyping}
\end{figure}

The consistent-subtyping rules are exactly the same as the subtyping rules,
except those that handle the dynamic type.

\begin{figure}[!ht]
\textbf{Consistent-Subtyping}\\
\dstart
\begin{footnotesize}
$$
\begin{array}{c}
\begin{array}{c}
\mylabel{C-LITERAL}\\
\senv \vdash L \lesssim L
\end{array}
\;
\begin{array}{c}
\mylabel{C-FALSE}\\
\senv \vdash \False \lesssim \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{C-TRUE}\\
\senv \vdash \True \lesssim \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-NUMBER}\\
\senv \vdash {<}{\it number}{>} \lesssim \Number
\end{array}
\;
\begin{array}{c}
\mylabel{C-STRING}\\
\senv \vdash {<}{\it string}{>} \lesssim \String
\end{array}
\;
\begin{array}{c}
\mylabel{C-BASE}\\
\senv \vdash B \lesssim B
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-NIL}\\
\senv \vdash \Nil \lesssim \Nil
\end{array}
\;
\begin{array}{c}
\mylabel{C-VALUE}\\
\senv \vdash T \lesssim \Value
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY1}\\
\senv \vdash T \lesssim \Any
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY2}\\
\senv \vdash \Any \lesssim T
\end{array}
\;
\begin{array}{c}
\mylabel{C-SELF}\\
\senv \vdash \Self \lesssim \Self
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-UNION1}\\
\dfrac{\senv \vdash T_{1} \lesssim T \;\;\;
       \senv \vdash T_{2} \lesssim T}
      {\senv \vdash T_{1} \cup T_{2} \lesssim T}
\end{array}
\;
\begin{array}{c}
\mylabel{C-UNION2}\\
\dfrac{\senv \vdash T \lesssim T_{1}}
      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-UNION3}\\
\dfrac{\senv \vdash T \lesssim T_{2}}
      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-FUNCTION}\\
\dfrac{\senv \vdash P_{2} \lesssim P_{1} \;\;\;
       \senv \vdash R_{1} \lesssim R_{2}}
      {\senv \vdash P_{1} \rightarrow R_{1} \lesssim P_{2} \rightarrow R_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash F_{j} \lesssim F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \lesssim \{F_{1}, ..., F_{n}\}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \lesssim F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash F \lesssim F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F\}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-FIELD1}\\
\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
       \senv \vdash T_{1} \lesssim T_{2} \;\;\;
       \senv \vdash T_{2} \lesssim T_{1}}
      {\senv \vdash K_{1}:T_{1} \lesssim K_{2}:T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-FIELD2}\\
\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
       \senv \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash \Const \; K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-FIELD3}\\
\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
       \senv \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-VARIABLE}\\
\dfrac{X_{1} \lesssim X_{2} \in \senv}
      {\senv \vdash X_{1} \lesssim X_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-RECURSIVE1}\\
\dfrac{\senv, X_{1} \lesssim X_{2} \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash \mu X_{1}.T_{1} \lesssim \mu X_{2}.T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-RECURSIVE2}\\
\dfrac{X \lesssim X \not\in \senv \;\;\;
      \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-RECURSIVE3}\\
\dfrac{X \lesssim X \in \senv \;\;\;
       \senv \vdash X \lesssim T_{2}}
      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-RECURSIVE4}\\
\dfrac{X \lesssim X \not\in \senv \;\;\;
       \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-RECURSIVE5}\\
\dfrac{X \lesssim X \in \senv \;\;\;
       \senv \vdash T_{1} \lesssim X}
      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-VOID}\\
\senv \vdash \Void \lesssim \Void
\end{array}
\;
\begin{array}{c}
\mylabel{C-VARARG}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \cup \Nil}
      {\senv \vdash T_{1}* \lesssim T_{2}*}
\end{array}
\;
\begin{array}{c}
\mylabel{C-TUPLE1}\\
\dfrac{\senv \vdash T_{1} \lesssim T_{2} \;\;\;
       \senv \vdash S_{1} \lesssim S_{2}}
      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2} \times S_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-TUPLE2}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \;\;\;
       \senv \vdash T_{1}* \lesssim S_{2}}
      {\senv \vdash T_{1}* \lesssim T_{2} \times S_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-TUPLE3}\\
\dfrac{\senv \vdash T_{1} \lesssim T_{2} \cup \Nil \;\;\;
       \senv \vdash S_{1} \lesssim T_{2}*}
      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2}*}
\end{array}
\end{array}
$$
\end{footnotesize}
\dend
\caption{The consistent-subtyping relation}
\label{fig:consistent_subtyping}
\end{figure}

\section{Related Lua projects}
\label{sec:related}
